---
title: "Investigation of Weather Events with Greatest Public Health and Economic Effects"
author: "Eric Oden"
date: "8/5/2021"
output: html_document
---
# Synopsis

Using the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm 
database, we seek to identify the weather events with the greatest effect on
public health and the economy. The NOAA database tracks major weather events in 
the United States from the years 1950 to 2011, and includes the number of 
injuries/fatalities and costs due to property/crop damage for each. Summing 
these quantities across the data set, we find tornadoes are by far the most
consequential weather events with regards to both public health and the economy.

# Data Processing

Import the relevant libraries:

```{r,message=FALSE}
library(dplyr)
library(reshape2)
library(ggplot2)
library(lubridate)
```

The data is loaded into **R** by downloading the publicly available [data set](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2), 
and converted into a data.frame object
```{r import, cache=TRUE}
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download.file(url = url, destfile = 'data/storm_data.csv')
df <- read.csv(file = 'data/storm_data.csv')
```

The data.frame has columns named EVTYPE, FATALITIES, and INJURIES 
corresponding to the weather event, the associated number of fatalities, and
associated number of injuries, respectively. There are also the columns PROPDMG,
PROPDMGEXP, CROPDMG, and CROPDMGEXP which give the costs due to property and 
crop damage. PROPDMG gives the mantissa, and PROPDMGEXP gives the exponent for 
the property damage costs due to the event in USD.

Since we are interested in identifying the weather events causing the most
health/economic damage, we filter out the rows with zero values in each column.

```{r subset, cache=TRUE}
df <- filter(df, PROPDMG > 0 | CROPDMG > 0 | FATALITIES > 0 | INJURIES > 0) 
```


To observe the public health effects, we create a new data.frame which 
contains the total fatalities and total injuries for each weather event. This is 
done using a sequence of **dplyr** functions which extract only the relevant 
columns (i.e., EVTYPE, FATALITIES, and INJURIES) from the original data.frame, 
group the rows by the weather event, and then sum over each group of rows. The 
new data.frame is then sorted by total injuries plus total fatalities in 
decreasing order.

```{r process_health, message=FALSE, cache=TRUE}
health_summary <- df %>% 
        select(EVTYPE, FATALITIES, INJURIES) %>% 
        group_by(EVTYPE) %>% 
        summarize(TOTAL_INJURIES = sum(INJURIES),
                  TOTAL_FATALITIES = sum(FATALITIES)) %>% 
        arrange(desc(TOTAL_FATALITIES + TOTAL_INJURIES))
```

To observe economic effects, the process is repeated, extracting instead the 
costs associated with property and crop damage.

We must calculate the true costs by using the information in the PROPDMG,
PROPDMGEXP, CROPDMG, and CROPDMGEXP columns.

```{r calc_costs}
key <- c("K" = 10^3, "M" = 10^6, "B" = 10^9)
df2 <- df %>% 
    select(EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP) %>%
    filter(PROPDMG > 0 | CROPDMG > 0) %>% 
    mutate(PROPCOST = as.numeric(PROPDMG * key[PROPDMGEXP])) %>% 
    mutate(CROPCOST = as.numeric(CROPDMG * key[CROPDMGEXP]))
```

```{r process_econ}
econ_summary <- df2 %>% 
    select(EVTYPE, PROPCOST, CROPCOST) %>% 
    group_by(EVTYPE) %>% 
    summarize(TOTAL_PROPCOST = sum(PROPCOST, na.rm = TRUE),
              TOTAL_CROPCOST = sum(CROPCOST, na.rm = TRUE)) %>% 
    arrange(desc(TOTAL_PROPCOST + TOTAL_CROPCOST))
```

# Results

We plot a clustered bar chart showing fatalities and injuries associated with 
the ten weather events with the greatest sum of the two variables.

```{r plot_health, message=FALSE}
long_health_summary <- melt(health_summary[0:10,], id.vars = "EVTYPE") 
ggplot(long_health_summary, 
       aes(x = reorder(EVTYPE, -value), y = value, fill=variable)) +
    geom_bar(stat='identity', position = "dodge") +
    xlab("Weather Event") +
    ylab("Total Injuries/Fatalities") +
    ggtitle("Weather Events Most Harmful to Public Health") +
    theme(axis.text.x = element_text(angle=45, hjust=1)) +
    theme(legend.title = element_blank()) +
    scale_fill_discrete(breaks=c("TOTAL_INJURIES", "TOTAL_FATALITIES"),
                        labels=c("Total Injuries", "Total Fatalities")) +
    labs(caption = "Total fatalities and Injuries for the ten weather events
         with the greatest sum of the two variables. We observe tornadoes
         cause the greatest number of injuries by far, while fatalities have 
         greater spread.")
```

We do the same for the economic effects, showing the property damage and crop 
damage for the ten weather events with the greatest sum of the two variables.

```{r plot_econ, message=FALSE}
long_econ_summary <- melt(econ_summary[0:10,], id.vars = "EVTYPE") 
ggplot(long_econ_summary, 
       aes(x = reorder(EVTYPE, -value), y = value/10^9, fill=variable)) +
    geom_bar(stat='identity', position = "dodge") +
    xlab("Weather Event") +
    ylab("Property/Crop Damage Costs (Billions USD)") +
    ggtitle("Weather Events with Greatest Economic Consequences") +
    theme(axis.text.x = element_text(angle=45, hjust=1)) +
    theme(legend.title = element_blank()) +
    scale_fill_discrete(breaks=c("TOTAL_PROPCOST", "TOTAL_CROPCOST"),
                        labels=c("Property Damage Costs", "Crop Damage Costs")) +
    labs(caption = "Total costs due to property and crop damage for the ten 
         weather events with the greatest sum of the two variables. We observe
         floods cause the most property damage.")
```

There is a clear indication that **tornadoes** have the greatest effect on 
public health, and **floods** have the greatest effect on the economy.
